Typed Interactions for NLU in Opal

### 2017-12-13 paper.tex
Cleaned up part of the formalism

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{subcaption}

@@

@@

In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
define a grammar of types, $\tau$ which will represent Opal types specifically.
We can think of $\tau$ as representing the final form of the data that a user
hopes to obtain, with no information about the entities that generated the data.
We add a few types that are noteworthy: the type {\bf string} is added to
account for FreeText entities; $\langle {\ell} \rangle$ represents a disjunction
of literals, and covers the case of Keywords entities.

@@

The \ff{lookup} function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free-text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities.
$$ \ff{Cfg} \equiv \ff{Tag} \rightharpoonup \tau $$
Now, we call a response well-formed with respect to $\ff{cfg} : \ff{Cfg}$ if and
only if it returns appropriate data on \ff{lookup}. More formally, for some $r :
\ff{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{\text{cfg}} \iff \\
  &\forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(\etag{t})\ \text{defined} \Rightarrow \ff{lookup}(\etag{t}, r) : \ff{cfg}(\etag{t})
\end{align*}

@@

We will define \ff{parse} by case analysis on the declaration.
\begin{align*}
  \ff{parse}(\text{FreeText}(\etag{t}), r) = \ff{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a {\bf string}, we can simply look up the value of the
appropriate entity from the response.
\begin{align*}
  \ff{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \ff{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from \ff{lookup}.
\begin{align*}
  \ff{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \ff{parseTy}(rec[\ff{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for \ff{parse} simply matches the value in the response to an
index into the record, and interprets that record entry using \ff{parseTy}.

@@

\subsection{Correctness}
Given a declaration $d$, we generate a configuration, \ff{cfg}. Then,
\begin{flalign*}
  &\forall\ (r : \textsf{Resp}).\\
  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
\end{flalign*}
Note that since \ff{parse} is a dependent function, it will already return a
value of the correct type. Thus, we only need to care that it is defined. Proof
of this fact is clear by induction on $d$.

>>>>

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{subcaption}

@@

\newtheorem{theorem}{Theorem}

@@

In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
define a grammar of types, $\tau$. We can think of $\tau$ as representing the
final form of the data that a user hopes to obtain, with no information about
the entities that generated the data. They map almost directly into types that
are used by the underlying Opal system. We add a few types that are noteworthy:
the type {\bf string} is added to account for FreeText entities; and $\langle
{\ell} \rangle$ represents a disjunction of literals, and covers the case of
Keywords entities.

@@

The \ff{lookup} function always returns a string, but the string might mean any
number of things. The data might be a particular literal, in the case of a
keywords or trait entity, or an entire phrase in the case of a free-text entity.
We define configurations as functions that map a subset of entity tags to the
type of data expected for those entities.
$$ \ff{Cfg} \equiv \ff{Tag} \rightharpoonup \tau $$
Now, we call a response well-formed with respect to $\ff{cfg} : \ff{Cfg}$ if and
only if it returns appropriate data on \ff{lookup}. More formally, for some $r :
\ff{Resp}$,
\begin{align*}
  &r\ \text{well-formed}_{\ff{cfg}} \iff \\
  &\forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(\etag{t})\ \text{defined} \Rightarrow \ff{lookup}(\etag{t}, r) : \ff{cfg}(\etag{t})
\end{align*}

When constructing a member of \ff{Cfg}, we must take care that the mapping
respects entity types. Thus, if \etag{t} is a tag for a free-text entity,
$\ff{cfg}(\etag{t})$ is allowed to be any string. If, instead, \etag{t} is a tag
for keywords or trait entities, $\ff{cfg}(\etag{t})$ should return a union of
literals that represents the appropriate data.

@@

In the event that $r$ is not well-formed with respect to the configuration
generated by $d$, $\ff{parse}(d, r)$ is undefined. Otherwise, we define
\ff{parse} by case analysis on the declaration.
\begin{align*}
  \ff{parse}(\text{FreeText}(\etag{t}), r) = \ff{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a {\bf string}, we can simply look up the value of the
appropriate entity from the response.
\begin{align*}
  \ff{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \ff{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from \ff{lookup}.
\begin{align*}
  \ff{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \ff{parseTy}(rec[\ff{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for \ff{parse} simply matches the value in the response to an
index into the record, and interprets that record entry using \ff{parseTy}.

@@

\subsection{Correctness}
We would like to bring both parts of the formalism together to show that our
system works the way we would like it to. The main property that we would like
to show is that \ff{parse} is correct.
\begin{theorem}[\ff{parse} Correct]
  Let $d$ be a declaration in $\fcy{D}$, and generate a configuration, \ff{cfg},
  from $d$. Then,
  \begin{flalign*}
    &\forall\ (r : \textsf{Resp}).\\
    &r\ \text{well-formed}_{\ff{cfg}} \Rightarrow \ff{parse}(d, r)\ \text{defined}
  \end{flalign*}
\end{theorem}
\begin{proof}
  By induction on $d$.
\end{proof}
Note that since \ff{parse} is a dependent function, it will already return a
value of the correct type. Thus, we only need to care that it is defined.
